---
title: "Getting started with the artemis package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with the artemis package}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(
  eval = FALSE, # for while editing vignette
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(artemis)
```
(Adapted from manuscript Espe et al, in review)

## Background

The `artemis` R package was created to aid in the design and analysis of
environmental DNA (eDNA) survey studies by offering a custom suite of
models for quantitative polymerase chain reaction (qPCR) data from extracted eDNA samples. Data from eDNA sampling surveys is often analyzed with occupancy models or GLMS, but there are
several characteristics of qPCR data in particular which made us feel that it would
benefit from a different modeling approach.  Specifically, our approach to this data makes use of Bayesian truncated latent variable models written in [Stan](https://mc-stan.org).


## Modeling qPCR eDNA Data with `artemis`

We created the `artemis` R package to implement Bayesian censored
latent variable models, which mitigate the weaknesses of common
statistical analysis techniques applied to qPCR data. Additionally,
`artemis` includes utilities to aid in the design and analysis of eDNA
survey studies, including simulation and power analysis functions.

At its core, `artemis` is a specialized Generalized Linear
Model, where the predictors are assumed to additively affect the
response variable, in this case $ln[eDNA]$, 

$$ ln[eDNA]_{i} = X_{i} \beta $$ 

where $\beta$ is a vector of effects on $ln[eDNA]_{i}$, and $X_{i}$
is a vector of predictors.  Since `artemis` directly models the
effect of the predictors on the latent (unobserved) variable, [eDNA],
it is unnecessary for the researcher to back-transform the data prior
to modeling. Internally, `artemis` conducts this conversion using the
user-supplied values for the formula,

$$\hat{Cq_i} = \alpha_{std\_curve} + \beta_{std\_curve}* ln[eDNA]_i  $$

Where $\alpha_{std\_curve}$ and $\beta_{std\_curve}$ are fixed values
from setting the standard curve in the lab prior to qPCR.  

Internally, the back-transformed $ln[eDNA]_i$ values are considered a
sample with measurement error from the true $ln[eDNA]_i$ value
($\hat{ln[eDNA]_i}$) in the extract. 

$$ ln[eDNA]_i \sim Trunc. Normal(\hat{ln[eDNA]_i}, \sigma_{Cq}, U) $$

Where the observed $ln[eDNA]_i$ values are censored at the
predetermined threshold, $U$. This threshold is back-transformed from
the threshold on Cq. Importantly, the $\hat{ln[eDNA]}$ values in
the model are not censored, allowing the latent variable to reflect the "true"
[eDNA] beyond the censorship point. The likelihood that a sampled $ln[eDNA]$
value will exceed the threshold is a function of the measurement error
and the estimated latent $\hat{ln[eDNA]_i}$ value. We calculate this
likelihood using the normal
cumulative distribution function, $\Phi()$,

$$ Pr[ln[eDNA]_i > U ] = 1 - \Phi(\hat{ln[eDNA]_i} - \mu_i / \sigma)$$

Thus, the models in `artemis` account for the data censoring process by
accounting for the probability that the observed value will exceed the
threshold. As detection limits vary with genetic assay, the upper
threshold on Cq in the model is adjustable by the user.

Lastly, there is an optional zero-inflated versions of both models,
`eDNA_lm_zinf()` and `eDNA_lmer_zinf()`. From multiple experiments, it
was observed there can be near-zero concentrations of eDNA even in
situations where higher concentrations were expected. This was
attributed to filter failures or other issues with sampling. To
account for this mechanism, the zero-inflated versions of the models
allow for zero [eDNA] concentrations from a secondary
mechanism. Currently, the functions do not support user-provided
predictors on the zero-inflated component, and just estimate a flat
probability of zero detections for all observations. However, users
can provide a prior for the expected probability of "true" zero
observations from a secondary mechanism.

This model formulation makes several assumptions:
 
  1. $ln[eDNA]$ is assumed to be uniform within a sample.
  
  2. $ln[eDNA]$ is sampled with normally distributed errors.
  	
  3. There are no false detections, i.e. the measurement error cannot
    result in a positive detection when eDNA is not present in the
    sample. 

Importantly, this formulation produces estimates of the effect sizes
which:

  - are modeled directly on $ln[eDNA]$ or copy number, rather than Cq, *therefore are independent of the standard curve and can be compared between studies that use different standard curves*. 
  
  - account for the data censoring at the upper limit of qPCR
    cycles, *which properly accounts for uncertainty and reduces bias in the estimates.*
	

## Installing `artemis` 

The `artemis` installation guide can be found [here](https://fishsciences.github.io/artemis/articles/artemis-installation-guide.html).


## A note on included sample data

Since `artemis` was originally written to analyze aquatic eDNA samples, the examples and sample data included in the package (see `?eDNA_data`) reference
variables that are specific to aquatic eDNA survey data. These variables include `volume`
(the the number of mililiters of water that went through the filter in
each sample) and `distance` (the distance in meters from the source of
eDNA).  For example, `eDNA_data` is from a "live car" experiment, where eDNA was sampled at known distances from a source of eDNA (fish in a net pen in unidirectional flow). 

In the following examples, `tech_rep` refers to qPCR replicates generated
 from a single extracted sample or filter; `rep` refers to a single extracted filter or sample.

## Overview of `artemis` Functionality

The functions included in `artemis` can be grouped into a few categories; each category is either introduced in this vignette, or addressed in detail by its own vignette.  [Available vignettes](https://fishsciences.github.io/artemis/articles/) are linked to individually below:

1. [Simulation of data](https://fishsciences.github.io/artemis/articles/simulation.html)
2. [Modeling data](https://fishsciences.github.io/artemis/articles/modeling.html)
3. Conducting post-hoc analyses
      * estimating probability of detection, generating predictions from model fits.
4. Conducting power and precision analyses

These categories are addressed in the sections that follow.    
<br>

The `artemis` package also includes methods for R's `plot()`, `summary()`, `data.frame()`, and `predict()` functions.


## Post-hoc analyses (*p(detect)*)

Often purpose of an eDNA sampling study is to inform a field sampling protocol. For these cases, we often want to know how likely we are to detect eDNA, given the way that we sampled. `est_p_detect()` returns the probability of getting a detection across all planned samples and replicates, where `n_rep` represents the product of the number of filters and technical replicates for each filter.  For example, if you planned to take 2-5 filters at each variable level and then complete 6 extractions for each filter, that would be `n_rep = 12:30`.

```{r eval=TRUE}
p_detect = est_p_detect(variable_levels = c(Intercept = 1, 
                                            Distance = 100),
                        betas = c(Intercept = -10.5, Distance = -0.03),
                        ln_eDNA_sd = 1, 
                        std_curve_alpha = 21.2, std_curve_beta = -1.5,
                        n_rep = 12:30)
p_detect
```

We can also plot this,

```{r fig = TRUE, eval=TRUE}
plot(p_detect)

```

Alternatively, we can also use a fit model to estimate the p(detect)
for a set of circumstances. Unlike the above, using a fit model allows
us to estimate the probability of detection with the uncertainty in
our effect size estimates taken into account.

```{r, eval=FALSE, fig = TRUE}
model_fit = eDNA_lm(Cq ~ Distance_m + Volume_mL, 
                    data = eDNA_data,
                    std_curve_alpha = 21.2, std_curve_beta = -1.5,
                    seed = 1234,
                    chains = 1)

# Intercept is automatically supplied if in model_fit
p_detect2 = est_p_detect(variable_levels = c(Distance_m = 100,
                                             Volume_mL = 50),
                        model_fit = model_fit, 
                        n_rep = 1:12)

print(p_detect2)

```

and we can plot the results,

```{r, fig = TRUE, eval = FALSE}
plot(p_detect2, error_width = 0.1)
```


<br>
<br>
---
<br>
<br>
