---
title: "ARTEMIS Installation Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{artemis-installation-guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
```

<br>
The easiest way to install `artemis` is from one of the pre-compiled binaries on the `artemis` [releases page](https://github.com/fishsciences/artemis/releases).  To do this, run the R code for your specific platform below:


```{r, eval=FALSE, include=TRUE}
# MacOSX
install.packages("https://github.com/fishsciences/artemis/releases/download/v0.9/artemis_0.9.0_R_OSX.tgz",
                 repos = NULL)

# LINUX
install.packages("https://github.com/fishsciences/artemis/releases/download/v0.9/artemis_0.9.0_R_x86_64-pc-linux-gnu.tar.gz", 
                 repos = NULL)

# Windows
install.packages("https://github.com/fishsciences/artemis/releases/download/v0.9/artemis.zip", 
                 repos = NULL)


```
<br>

### Installing `artemis` from source

Installing `artemis` from source on Windows is not currently well-supported; we recommend installing from the pre-compiled binary if you're on Windows.  If you're on MacOS or Linux and you prefer to install from source, then go ahead and do that with your function/utility of choice (`devtools::install_github()`, `utils::install.packages(type = "source")`, `R CMD INSTALL`, etc.).  
<br>
If you have sub-architecture you're really in to customizing, the source code is [here](https://github.com/fishsciences/artemis/releases), go nuts.

<br>
