---
title: "ARTEMIS Installation Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{artemis-installation-guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
```

<br>
The easiest way to install `artemis` is from one of the pre-compiled binaries on the `artemis` [releases page](https://github.com/fishsciences/artemis/releases).  To do this, run the R code for your specific platform below:


```{r, eval=FALSE, include=TRUE}
# MacOSX
install.packages("https://github.com/fishsciences/artemis/releases/download/v0.9/artemis_0.9.0_R_OSX.tgz",
                 repos = NULL)

# LINUX
install.packages("https://github.com/fishsciences/artemis/releases/download/v0.9/artemis_0.9.0_R_x86_64-pc-linux-gnu.tar.gz", 
                 repos = NULL)

# Windows
install.packages("https://github.com/fishsciences/artemis/releases/download/v0.9/artemis.zip", 
                 repos = NULL)


```

<br>

If you're set up to build compiled code from source on your machine, you can use also the `devtools` package to build `artemis` from its source on GitHub.  If you don't know whether your system is set up to do this, try running `devtools::has_devel()`.  

<br>

**If you see the message "`Your system is ready to build packages!`", and you're on Windows, we still recommend installing from the pre-compiled binary.**  If you're on MacOS or Linux, then go ahead and install from source however you like to do that (the `install_github()` function from the `devtools` package, `utils::install.packages("path-to-.tz", type = "source")`, `R CMD INSTALL`, etc.  


## MacOS or Linux from source using `devtools::install_github()`

```{r, eval=FALSE}
devtools::install_github("fishsciences/artemis", build_vignettes = FALSE) # you can change to TRUE if you want to launch the vignettes from R, but note that install will be slower.
```

If you have sub-architecture you're really in to customizing on install, the source code is [here](https://github.com/fishsciences/artemis/releases), go nuts.

<br>
