---
title: "artemis Installation Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ARTEMIS Installation Guide}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE
)
```

<br>
The easiest way to install `artemis` is from [CRAN](https://cran.r-project.org/package=artemis) with the `install.packages` function:


```{r, eval=FALSE, include=TRUE}

install.packages("artemis")

```
<br>

## Testing your installation

If your installation of `artemis` and its dependencies was successful, the following code should run without error (although you may see warning messages from `rstan` about Bulk/Tail Effective Samples Sizes being too low). If the first or second model returns an error that seems to have something to do with your `c++` compiler, you may need to [follow instructions to edit your `Makevars` or `Makevars.win` file](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started).

```{r, eval=FALSE}
library(artemis)

model_fit = eDNA_lm(Cq ~ scale(Distance_m) + scale(Volume_mL), 
                    data = eDNA_data,
                    std_curve_alpha = 21.2, std_curve_beta = -1.5)

model_fit2 = eDNA_lmer(Cq ~ scale(Distance_m) + scale(Volume_mL) + (1|FilterID),
                       eDNA_data,
                       std_curve_alpha = 21.2, std_curve_beta = -1.5)

```


## Installing `artemis` from source

Installing `artemis` from source on Windows is not currently well-supported; we recommend installing from the pre-compiled binary if you're on Windows.  If you're on MacOS or Linux and you prefer to install from source, then go ahead and do that with your function/utility of choice (`devtools::install_github()`, `utils::install.packages(type = "source")`, `R CMD INSTALL`, etc.).  
<br>
If you have sub-architecture you're really in to customizing, the source code is [here](https://github.com/fishsciences/artemis), go nuts.

<br>
