---
title: "Getting started with the artemis package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting_started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(
  eval = FALSE, # for while editing vignette
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library(artemis)
```

The `artemis` package was created to aid in the design and analysis of eDNA survey studies by offering a custom suite of models for eDNA sampling and qPCR data. There are several reasons we felt that modeling of eDNA data could benefit from this treatment:

  1. In eDNA samples, the concentration of eDNA is not directly measured. Instead, the amount of
     eDNA present is represented as a function of the number of quantification cycles of qPCR (hereafter referred to as the "Cq" value) completed before amplification takes place. eDNA concentration ($[eDNA]$) is then related to Cq via a standard curve generated in the lab for the target species. 
	 
  2. A lower Cq value corresponds to a higher concentration of eDNA in a
     sample. Above a pre-determined threshold, additional quantification cycles are not
     attempted. Therefore, "non-detection" is taken to be any sample
     which requires more than the threshold number of cycles to detect.  This data truncation in the observed response increases uncertainty in our estimate of the underlying response variable,
     eDNA concentration.
  
  3. The potential sources of measurement error in the extraction and qPCR processes are difficult to separate and quantify. For example, Cq values produced by qPCR become more variable at the threshold of detection, i.e. as the number of eDNA molecules available for amplification approaches zero.  This source of variability in the response is different from that produced by error introduced in the pipetting process during extraction, but they have the same effect on Cq (namely, increasing variability).
	 
The `artemis` package addresses these issues by directly modeling the latent (unobserved) eDNA
concentration, linking it to the truncated observed response via a standard curve, and estimating the measurement error in the process. The general model is as follows:

$$ Cq_i \sim Normal(\hat{Cq_i}, \sigma_{Cq}) \\
\hat{Cq_i} = \alpha_{std\_curve} + \beta_{std\_curve}* log[eDNA]_i  \\
log[eDNA]_{i} = \beta X_{i}$$

where $\beta$ is a vector of effects on $log[eDNA]$, and $\alpha_{std\_curve}$ and $\beta_{std\_curve}$ are fixed values provided by the standard curve.

    
    
As detection limits vary with genetic assay, the upper threshold on Cq in the model is adjustable.  Any values of $\hat{Cq_i}$ which are greater than the upper limit on $Cq_i$ are recorded as the threshold value.  For example, a $\hat{Cq_i}$ value of 42 is recorded as 40 when the upper limit is 40 cycles. 
   
   
This model formulation makes several assumptions:
 
  - $log[eDNA]$ is assumed to be uniform within a sample
  
  - $log[eDNA]$ is sampled without error.
  
  - All measurement error is introduced in the extraction/qPCR stage. Field sampling is assumed to take place without error.*
	
  - There are no false detections, i.e. the measurement error cannot
    result in a positive detection when eDNA is not present in the
    sample. 
  
*`*`Future versions of the artemis package might allow for measurement
error in both the field collection and qPCR stages.*

In `artemis`, the model is specified using a model formula, similar to
the `lm()` or `lmer()` functions. This model formula is used to
construct the model on $log[eDNA]$.

# A note on survey data

Because `artemis` was originally written to analyze aquatic eDNA samples, the examples and sample data included in the package have variables that are specific to aquatic eDNA survey data, and even more specifically to "live car" experiments, where eDNA is sampled at known distances from a source of eDNA (fish in a net pen in unidirectional flow, for example).  These variables include `volume` (the the number of mililiters of water that went through the filter in each sample) and `distance` (the distance in meters from the source of eDNA).  However, the functions in `artemis` do generalize to any eDNA survey data, so long as that study generates Cq values and is associated with a standard curve for the target species.

In the following examples, `tech_rep` refers to technical replicates extracted from a single eDNA sample; `rep` refers to a single eDNA sample.

# Simulations

One of the most basic tasks for the `artemis` package is simulating
data. Simulated data is useful for testing assumptions, designing a
study, or doing a post-hoc analysis. 

To simulate data, you first must provide a set of levels for each variable in your survey data.  In the following example, our variables are `distance`, `volume`, `tech_rep`, and `rep`.

```{r}
vars = list(Intercept = 1,
            distance = c(0, 15, 50),
            volume = c(25, 50),
            tech_rep = 1:10,
            rep = 1:3, Cq = 1)
```

Notice that we also provided a dummy variable for the intercept and response - this is not used
except internally to create the model matrix.  By default, the simulation functions construct a model matrix assuming a complete block design (i.e. every level of each variable is represented). The simulation function expands the variable list using `base::expand.grid()`. 

Second, you must provide a set of effect sizes for the variables you
provide. For example, if we were only simulating data using distance,
we might specify the effect sizes as:

```{r}
betas = c(intercept = -10.6, distance = -0.05, volume = 0.1)
```

It is important to note that these are the effect sizes on $log[eDNA]$, *not* on the Cq values themselves. All the parameters of the model are on the log scale, including the intercept. Hence, we specified the intercept to be -10, which corresponds to exp(-10)$[eDNA]$. In practice, if the default value for log[eDNA] is close to 0, some negative number will work for the intercept beta.

To simulate data from a fixed-effects model, we use the `sim_eDNA_lm()` function:

```{r warning = FALSE}
ans = sim_eDNA_lm(Cq ~ distance + volume, 
                  vars,
                  betas = c(intercept = -10.6, distance = -0.05, volume = 0.1),
                  sigma_Cq = 1, 
                  std_curve_alpha = 21.2, 
                  std_curve_beta = -1.5)
```

Notice that we must provide the parameters for the conversion formula
between $log[eDNA]$ and Cq values to the simulation
function, as well as the measurement error on Cq (`sigma_Cq`).

By default, only one instance of simulated data is created, but we can
easily simulate many data sets by adjusting the `n_sim` parameter,
e.g.

```{r eval = FALSE}
ans2 = sim_eDNA_lm(Cq ~ distance + volume, 
                  vars,
                  betas = c(intercept = -10.6, distance = -0.05, volume = 0.1),
                  sigma_Cq = 1, 
                  std_curve_alpha = 21.2, 
                  std_curve_beta = -1.5,
                  n_sim = 500)
```

If we want to simulate data for an incomplete block design, we can specify our own variable data.frame, `X`, in which case the variable levels are ignored and the provided data.frame is used to
create a model.matrix.

To simulate for a mixed-effects model, we use the function
`sim_eDNA_lmer()`. Random effects are specified using the syntax of
the `lme4` package, e.g. `(1|random effect)`:

```{r warning = FALSE}
ans3 = sim_eDNA_lmer(Cq ~ distance + volume + (1|rep) + (1|tech_rep),
                     vars,
                     betas = c(intercept = -10.6, distance = -0.05, volume = 0.01),
                     sigma_Cq = 1,
                     sigma_rand = c(0.1, 0.1), #stdev of the random effects
                     std_curve_alpha = 21.2,
                     std_curve_beta = -1.5)

```

When we are simulating data with random effects, we must provide the
standard deviation of the random effects (one for each effect) in addition to the information for the fixed effects. Random effects are generated for these simulations using this standard deviation (`sigma_rand`). When `n_sims` > 1, different random effects are generated for each simulation.

The results of a simulation can be inspected via the `summary` and `plot`
methods. These can be helpful when assessing how realistic the simulated
data are, and to test the assumptions behind the model and the betas
used. For example, the results simulated for the fixed-effects model above can be summarized by calling:

```{r}
summary(ans)
```

The default behavior of `summary()` is to give the marginal distribution of the simulated data,
i.e. the effects summarized by each level of the variables used to
simulate the data. *WARNING: these marginal effects can be misleading
in cases where interactions are used in the formula.* The summary also
includes an empirical summary of the percent of each level which was
detected, i.e. had a Cq value below the threshold.

So in the output above, 95% of the simulated data with distance = 0 was between  27.8 and 31.6.
The p_detect column is the percentage of each effect that was below the detection threshold.  In this simulation, the intercept is pretty low and the effects of distance and volume are small.  100% of all levels resulted in samples below the detection threshold.

**TODO** Plot methods

## Study design

Simulations are helpful for designing new studies. Given a set of
assumptions about the effect sizes, we can simulate how well we are
able to either detect eDNA or how accurately we can estimate those
effects given a number of replications and variable levels in the design.

For example, suppose we wanted to design a study exploring the effects
of different distances on the detection of eDNA. From previous
studies, we have estimated the effect of distance to be -0.04 decrease
in log(eDNA concentration) per meter distance. If we want to have a
sense of how many replicates we need for this study to have at least
80% power (at least 80% chance to estimate the effect of distance
without including 0 in the CI), we can get a quick estimate of this
using the `est_power_lm()` and a given number of samples, say 5 reps. To
do this, we specify the number of replicates in the variable_list as
`rep = 1:5`, and specify our effect sizes,

```{r warning = FALSE}
dist_power = est_power_lm(Cq ~ 1 + distance,
                          variable_list = list(Cq = 1, distance = c(0,100,300), rep = 1:5),
                          betas = c(intercept = -10.6, distance = -0.04),
                          sigma_Cq = 1, 
                          std_curve_alpha = 21.2, std_curve_beta = -1.5,
                          type = "exclude_zero")

dist_power

```

It appears that 5 samples was not enough in this case. We could
keep trying different number of replicates until we get the number we
would like. However, it is easier to use the function
`est_power_range_lm`, which will automate this process for us. The
call is similar to the above, but instead of specifying the number of
replicates in the `variable_list`, we provide a `rep_range`, which is
`seq(2,20,2)` (every 2 from 2 to 20) by default

```{r warning = FALSE}
dist_power_range = est_power_range_lm(Cq ~ 1 + distance,
                                variable_list = list(Cq = 1, distance = c(0,100,300)),
                                betas = c(intercept = -10.6, distance = -0.04),
                                sigma_Cq = 1, 
                                std_curve_alpha = 21.2, std_curve_beta = -1.5,
                                type = "exclude_zero")

dist_power_range
```

For this study, it looks like we will need greater then 8 replicates
to have our desired 80% power. 

# Modeling

The second purpose of the artemis package is to fit a model to
collected data. Similar to the simulation functions, the artemis
package has two functions which fit a Bayesian latent variable model
to eDNA data; `eDNA_lm()` for fixed effects models and `eDNA_lmer()`
for mixed effects models.

The syntax of these functions is similar to the simulation
functions. We provide a model formula, a data.frame with the observed
data, and then the parameters for standard curve formula to convert
between log(eDNA concentration) and Cq values. For example, to fit a
model to the sample data included in artemis, `eDNA_samples`:

```{r warning = FALSE}
model_fit = eDNA_lm(Cq ~ Distance, eDNA_samples,
                    std_curve_alpha = 21.2, std_curve_beta = -1.5)

```

The modeling functions use [Stan](mc-stan.org) to fit the Bayesian
model. Full control of the MCMC algorithm can be accomplished by
passing arguments to `rstan::stan()` in the modeling functions. For
example, 


```{r warning = FALSE}
model_fit = eDNA_lm(Cq ~ Distance, eDNA_samples,
                    std_curve_alpha = 21.2, std_curve_beta = -1.5,
                    iter = 1000)

```

By default, `artemis` suppresses the (often verbose) output from Stan,
but users can (and should) enable it when the model is slow or the
output suggests the MCMC algorithm might not have converged, for
example,

```{r warning = FALSE}
model_fit = eDNA_lm(Cq ~ Distance, eDNA_samples,
                    std_curve_alpha = 21.2, std_curve_beta = -1.5,
                    iter = 100, n_chain = 1,
                    verbose = TRUE)
```

To fit a model with a random effect, use the
`eDNA_lmer()` function and specify the random effects using the syntax
used by `lme4`:
```{r warning = FALSE}
model_fit2 = eDNA_lmer(Cq ~ Distance + Volume + (1|TechnicalRep),
                       eDNA_samples,
                       std_curve_alpha = 21.2, std_curve_beta = -1.5)

```

As with the simulations, the model results can be summarized with
`summary()` and can be plotted with `plot()`,

```{r}
summary(model_fit2)
```

**TODO:** PLOT

## Post-hoc p(detect)

It is common that the purpose of a eDNA sampling study is to inform
the deployment of an automated sampler in the field. For these cases,
we often want to know how likely we are to detect eDNA given the
estimated parameters of the model and a set of conditions. For these
circumstances, we can use the `est_p_detect()` function to see how
well we can detect given effect sizes and conditions,

```{r}
p_detect = est_p_detect(variable_levels = c(Intercept = 1, Distance = 100),
                        betas = c(Intercept = -10.5, Distance = -0.04),
                        Cq_sd = 1, std_curve_alpha = 21.2, std_curve_beta = -1.5,
                        n_rep = 1:12)
print(p_detect)
```
and we can plot this,

```{r fig = TRUE}
plot(p_detect)

```

Alternatively, we can also use a fit model to estimate the p(detect)
for a set of circumstances. Unlike the above, using a fit model allows
us to estimate the probability of detection with the uncertainty in
our effect size estimates taken into account.

```{r}
p_detect2 = est_p_detect(variable_levels = c(Intercept = 1, Distance = 100),
                        model_fit = model_fit, 
                        Cq_sd = 1, std_curve_alpha = 21.2, std_curve_beta = -1.5,
                        n_rep = 1:12)

print(p_detect2)
```

and we can plot the results,

```{r fig = TRUE}
plot(p_detect2)
```

Notice how we get uncertainty intervals when we use a fit model. 


