---
title: "Models included in the `artemis` package and when to use them"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling Types in `artemis`}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, warning = FALSE}
library(artemis)
compile_models(rewrite = FALSE)
```

- `eDNA_lm()`: Fixed effects censored data model with Normal errors. Use when threshold of detection is greater than single copies of DNA and there are no grouping variables (random effects)

- `eDNA_lmer()`: Mixed effects censored data model with Normal errors. Same as above, but there are grouping variables (random effects).

- `eDNA_zinf_lm/lmer()`: Same as above, but the response is zero-inflated (has more zero values than expected under the normal model assumptions).

- `eDNA_count_lm/lmer()`: Use when the threshold of detection is single copies of DNA. Requires the response to be converted to copy number (integer values). 


## Which model should I use?

```mermaid
graph TD;
  A(Start) --> B{"Does your data contain 0 (CT = 40) values?"};
  B -->|Yes| C{"Do your data represent single-digit copy numbers?"};
  B -->|No| D("Use a standard model from another package (e.g. glm(), lm(), etc.)");
  C -->|Yes| E("Use the Poisson-distributed GLM models (eDNA_count())");
  C -->|No| F{"Do the zeros in the data occur in otherwise high concentrations?"};
  F -->|Yes| G("Use a zero-inflated model (eDNA_lm*_zinf())");
  F -->|No| H("Use a censored-data model (eDNA_lm*())"); 
```
