---
title: "Modeling eDNA qPCR Data with `artemis`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling eDNA qPCR Data with `artemis`}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, warning = FALSE}
library(artemis)
```

# Introduction

A primary purpose of the `artemis` package is to facilitate modeling of qPCR data from eDNA samples.  It does this via two functions: `eDNA_lm()` for fixed effects models and `eDNA_lmer()` for mixed effects models.

The underlying [Stan](https://mc-stan.org) models are compiled on install. Thereafter, the models will not need
to be re-compiled. The model's Stan code can be found in the `artemis` source code [here](https://github.com/fishsciences/artemis/tree/main/inst/stan). 

## Model Inputs

Both modeling functions require the following inputs:

1. A vector of numeric Cq values (one for each qPCR replicate). Cq values corresponding to non-detections for your assay should be recorded as the threshold value (the default is 40.0 cycles).

2. The intercept value $\alpha$ and the slope value $\beta$ from a
   standard curve equation associated with the qPCR analysis. This is
   used to convert the observed Cq values to the corresponding log
   concentration of eDNA.

An example of qPCR data in the correct format for modeling with `artemis` can be viewed by calling `eDNA_data`, which is a `data.frame` with Cq values from live car experiments completed in the California Sacramento-San Joaquin Delta with Delta Smelt:

```{r}
head(eDNA_data)
str(eDNA_data)
```


Note that there are no variable levels with missing or `NA` values - Stan models do not take `NA` values, and any rows with `NA`s in the data will be dropped in the construction of the model matrix when the data is prepped for modeling.

<br>

# Fixed effects models with `eDNA_lm()`

To fit a fixed effects model to the sample `eDNA_data` where `Distance_m` is the only predictor, we give the function a model formula and the input data listed above:

```{r warning = FALSE, eval=FALSE}

model_fit = eDNA_lm(Cq ~ Distance_m, 
                    data = eDNA_data,
                    std_curve_alpha = 21.2, std_curve_beta = -1.5)

```

<br>

The model functions, similar to `lm()` in base R, will automatically
add an intercept term. You can explicitly omit the intercept if you
have a good reason for doing so.  Full control of the MCMC algorithm
can be accomplished by adding these control arguments to the end of
the `eDNA_lm*()` call, which then passes them on to
`cmdstanr::model$sample()`. *Note: the arguments for the model follow
the naming conventions for `cmdstan`, which differ from those used by
`rstan`. Documentation of these arguments can be found
[here](https://mc-stan.org/cmdstanr/reference/model-method-sample.html)*

For example,


```{r warning = FALSE, message=FALSE, eval=FALSE}
model_fit = eDNA_lm(Cq ~ Distance_m, 
                    data = eDNA_data,
                    std_curve_alpha = 21.2, std_curve_beta = -1.5,
                    seed = 1234, 
                    chains = 1) # we don't recommend sampling just 1 chain; the default is 4

```

<br>

<!-- By default, `artemis` suppresses the (often verbose) output from Stan, -->
<!-- but users can (and should) enable it with `verbose = TRUE` when the model is sampling slowly or the -->
<!-- output suggests the MCMC algorithm might not have converged, for example, -->

<!-- ```{r warning = FALSE, eval=TRUE} -->
<!-- model_fit = eDNA_lm(Cq ~ Distance + Volume + Distance*Volume,  -->
<!--                     data = eDNA_data, -->
<!--                     std_curve_alpha = 21.2, std_curve_beta = -1.5, -->
<!--                     iter = 500,  -->
<!--                     seed = 1234,  -->
<!--                     n_chain = 1, # set to 1 for vignette; default is 4 -->
<!--                     verbose = TRUE) -->
<!-- ``` -->

<!-- <br> -->

# Mixed effects models with `eDNA_lmer()`

<br>

To fit a model with one or more random effect(s), use the `eDNA_lmer()` function. Random effects are specified using the same syntax as the `lme4` package, e.g. `(1|random effect)`:

```{r warning = FALSE, eval=FALSE}
d = eDNA_data # create a copy to modify 
d$Year = factor(sample(2018:2020, size = nrow(d), replace = TRUE)) # create a random variable

model_fit2 = eDNA_lmer(Cq ~ Distance_m + Volume_mL + (1|Year),
                       data = d,
                       std_curve_alpha = 21.2, std_curve_beta = -1.5,
                       seed = 1234) 

```

# Summarizing and plotting model output

As with the simulation objects, the model results can be summarized or plotted with default methods using
`summary()` and `plot()`, or converted to a `data.frame` object for further manipulation.

```{r, eval=FALSE}
summary(model_fit)

plot(model_fit, pars = c("intercept", "betas"))
```

<br>

Matching `lme4` convention, random effects are not included in the default `summary()` output.  You can view or plot the random effects estimates by subsetting the `stanfit` slot of the model object with `@`, and specifying the `random_betas` parameters with the `pars` argument:

```{r, eval=FALSE}
rstan::summary(model_fit2@stanfit, pars = "rand_betas", probs = c(0.50, 0.025, 0.975))$summary
plot(model_fit2, pars = "rand_betas")
```

Similar to `lme4`, there is a `ranef` method for `eDNA_model` objects

```{r eval=FALSE}
ranef(model_fit2)
```

<br>

# Further notes on modeling

Because the models implemented in `artemis` are Bayesian, you will get the most out of their results when you can work with and summarize posterior probabilities.  Some helpful resources for this are the [Stan User's Guide](https://mc-stan.org/docs/2_22/stan-users-guide/index.html), and the [stanfit objects vignette](http://mc-stan.org/rstan/articles/stanfit_objects.html) from the `rstan` package.

<br>
---
<br>
