---
title: "Modeling eDNA qPCR Data with `artemis`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling eDNA qPCR Data with `artemis`}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, warning = FALSE}
library(artemis)
```

(Adapted from manuscript Espe et al, in review)

# Introduction

A primary purpose of the `artemis` package is to facilitate modeling of qPCR data from eDNA samples.  It does this via two modeling functions: `eDNA_lm()` for fixed effects models and `eDNA_lmer()` for mixed effects models. These functions mirror the semantics of R's built in `lm()` and lme4's `lmer()`.

## Model Inputs

Both modeling functions require the following inputs from the data:

1. A vector of numeric Cq values (quantification cycles), one for each qPCR replicate. Cq values corresponding to non-detections for your assay should be recorded as the threshold value (the default is 40.0 cycles).

2. The intercept value $\alpha$ and the slope value $\beta$ from a
   standard curve equation associated with the qPCR analysis. This is
   used to convert the observed Cq values to the corresponding log
   concentration of eDNA. This conversion occurs internally.
   
3. A threshold value of the most cycles which are attempted in qPCR (defaults to 40 cycles).

An example of qPCR data in the correct format for modeling with `artemis` can be viewed by calling `eDNA_data`, which is a `data.frame` with Cq values from live car experiments completed in the California Sacramento-San Joaquin Delta with Delta Smelt:

```{r}
head(eDNA_data)
str(eDNA_data)
```

Note that there are no variable levels with missing or `NA` values in
these example data. However if there were `NA` values in the input
data set, any rows with `NA`s in the data will be dropped when the
data is prepped for modeling.  This is because Stan models cannot not
take `NA` values. Although `NA` values will be automatically dropped
from the data prior to modeling, we recommend removing `NA` values as
a separate step prior to modeling. This allows inspection and
potentially correction of the rows with `NA` values. For example,

```{r eval = FALSE}
na_vals = !complete.cases(eDNA_data)
eDNA_data[na_vals,] # visual inspection
```

<br>

# Fixed effects models with `eDNA_lm()`

Fixed effects models are primarily used with completely randomized
experiments without blocking variables. For most observational data or
blocked experimental data, mixed effects models are likely more
appropriate.

To fit a fixed effects model to the sample `eDNA_data` where `Distance_m` is the only predictor, we give the function a model formula and the input data listed above:

```{r warning = FALSE, eval=FALSE}

model_fit = eDNA_lm(Cq ~ Distance_m, 
                    data = eDNA_data,
                    std_curve_alpha = 21.2, std_curve_beta = -1.5)

```

Notice that we provide the standard curve parameters
(`std_curve_alpha` and `std_curve_beta` as separate arguments to the
function. In cases where there are multiple standard curve parameters
in use in the same dataset (e.g. using data from multiple labs or
experiments), the standard curve parameters can each be given as
vectors. These vectors must be the same length as the number of rows
in the data. 

<br>

The model functions, similar to `lm()` in base R, will automatically
add an intercept term. You can explicitly omit the intercept if you
have a good reason for doing so. Please see `?lm` for a more full
description of how to specify linear models in R.

Full control of the MCMC algorithm can be accomplished by adding these
control arguments to the end of the `eDNA_lm*()` call, which then
passes them on to `rstan::sampling()`. Available arguments for MCMC
control can be found in the help for `rstan::sampling`.

For example,


```{r warning = FALSE, message=FALSE, eval=FALSE}
model_fit = eDNA_lm(Cq ~ Distance_m, 
                    data = eDNA_data,
                    std_curve_alpha = 21.2, std_curve_beta = -1.5,
                    seed = 1234, 
                    chains = 1) # we don't recommend sampling just 1 chain; the default is 4

```

<br>

# Mixed effects models with `eDNA_lmer()`

<br>

Random or mixed effects models are typically used when there are
grouping factors which need to be accounted for in the model
(e.g. blocking variables, subsamplings from a single filter, etc.). 

To fit a model with one or more random effect(s), use the `eDNA_lmer()` function. Random effects are specified using the same syntax as the `lme4` package, e.g. `(1|random effect)`.

For example, to specify a random effect for "Year",

```{r warning = FALSE, eval=FALSE}
d = eDNA_data # create a copy to modify 
d$Year = factor(sample(2018:2020, size = nrow(d), replace = TRUE)) # create a random variable

model_fit2 = eDNA_lmer(Cq ~ Distance_m + Volume_mL + (1|Year),
                       data = d,
                       std_curve_alpha = 21.2, std_curve_beta = -1.5,
                       seed = 1234) 

```

# Summarizing and plotting model output

As with the simulation objects, the model results can be summarized or plotted with default methods using
`summary()` and `plot()`, or converted to a `data.frame` object for further manipulation.

```{r, eval=FALSE}
summary(model_fit)

plot(model_fit, pars = c("intercept", "betas"))
```

Additional arguments can be provided to the plot method, which are
passed to `rstan::plot` methods for `stanfit` objects. More details
are available via `?rstan::plot`.

<br>

Matching `lme4` convention, random effects are not included in the
default `summary()` output.  You can view a summary of the random
effects with `ranef()`,


```{r eval=FALSE}
ranef(model_fit2)
```

or by subsetting the `stanfit` slot of the model object with `@`, and
specifying the `random_betas` parameters with the `pars` argument:

```{r, eval=FALSE}
rstan::summary(model_fit2@stanfit, pars = "rand_betas", probs = c(0.50, 0.025, 0.975))$summary
plot(model_fit2, pars = "rand_betas")
```

<br>

# Further notes on modeling

Because the models implemented in `artemis` are Bayesian, you will get the most out of their results when you can work with and summarize posterior probabilities.  Some helpful resources for this are the [Stan User's Guide](https://mc-stan.org/docs/2_22/stan-users-guide/index.html), and the [stanfit objects vignette](http://mc-stan.org/rstan/articles/stanfit_objects.html) from the `rstan` package.

## Useful modeling advice

This is a collection of advice for modeling eDNA data with the
`artemis` package.

1. Center and scale your predictor values: `artemis` uses MCMC to
   estimate values, and this will be more efficient if the predictor
   values are not on vastly different scales. In general, the MCMC
   will be the most efficient when the predictors are roughly centered
   at 0, and have stdev of 1.
   
2. Use priors: The default priors in `artemis` follow the conventions
   of the `rstanarm` package, and are weakly informative. When the
   data do not strongly inform the parameter estimates, the model fit
   can be improved by specifying stronger priors. 

## Other information

The underlying [Stan](https://mc-stan.org) models are compiled on install. Thereafter, the models will not need
to be re-compiled. The model's Stan code can be found in the `artemis` source code [here](https://github.com/fishsciences/artemis/tree/main/inst/stan). 


<br>
---
<br>
