---
title: "Modeling eDNA qPCR Data with `artemis`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling eDNA qPCR Data with `artemis`}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE, warning = FALSE}
library(artemis)
```

# Introduction

A primary purpose of the `artemis` package is to allow fitting Bayesian latent variable models to
collected qPCR data. Similar to the simulation functions, the `artemis`
package has two functions with which to do this: `eDNA_lm()` for fixed effects models and `eDNA_lmer()`
for mixed effects models.


# Fitting fixed effects models with `eDNA_lm()`

This function takes a model formula (with syntax identical to that of the `lme4` package), a `dataframe` or `list` of the observed data, and the parameters for the data's associated standard curve equation. For example, to fit a
model to the sample data included in `artemis`, `eDNA_data`:

```{r warning = FALSE, eval=FALSE}

model_fit = eDNA_lm(Cq ~ Distance, 
                    data = eDNA_data,
                    std_curve_alpha = 21.2, std_curve_beta = -1.5)

```

The modeling functions use [Stan](mc-stan.org) to fit the Bayesian
model. Full control of the MCMC algorithm can be accomplished by adding them to the end of the `eDNA_lm*()` call, which then passes the arguments to `rstan::stan()`. For example, 


```{r warning = FALSE, message=FALSE, eval=FALSE}
model_fit = eDNA_lm(Cq ~ Distance, 
                    data = eDNA_data,
                    std_curve_alpha = 21.2, std_curve_beta = -1.5,
                    iter = 1000, seed = 1234, n_chain = 2)

```

By default, `artemis` suppresses the (often verbose) output from Stan,
but users can (and should) enable it with `verbose = TRUE` when the model is slow or the
output suggests the MCMC algorithm might not have converged, for example,

```{r warning = FALSE, eval=TRUE}
model_fit = eDNA_lm(Cq ~ Distance + Volume + Distance*Volume, 
                    data = eDNA_data,
                    std_curve_alpha = 21.2, std_curve_beta = -1.5,
                    iter = 300, seed = 1234, n_chain = 2,
                    verbose = TRUE)
```

# Fitting mixed effects models with `eDNA_lmer()`

To fit a model with a random effect, use the
`eDNA_lmer()` function. Random effects are specified using the syntax of the `lme4` package, e.g. `(1|random effect)`:

```{r warning = FALSE, eval=FALSE}

model_fit2 = eDNA_lmer(Cq ~ Distance + Volume + (1|SampleID),
                       data = eDNA_data,
                       std_curve_alpha = 21.2, std_curve_beta = -1.5)

```

# Summarizing and plotting model output

As with the simulation objects, the model results can be summarized or plotted with default methods using
`summary()` and `plot()`,

```{r, eval=TRUE}
summary(model_fit)

plot(model_fit, pars = c(sprintf("betas[%d]", 2:4)))
```

